---
title: "Analysis of school problem dataset"
output: html_document
---

```{r global_options, include=FALSE}
library(car)
library(plyr)
library(dplyr)
library(ggplot2)
library(Rmisc)


```

This dataset contains questionnaire responses of 573 students from regular and hospital schools. 

We first load the cleaned data and compute the SDQ scores. To this end, we use a custom adaptation of the "official" R function provided on the [SDQ website](http://www.sdqinfo.com/c9.html).

```{r}
# import the data
data = read.csv("D:/lab/projects/schoolproblems2/data/schooldata.csv")

# import the function to compute sdq scores
source("D:/lab/projects/schoolproblems2/code/sdq_eval.r")

# compute sdq scores and append to dataframe
sdq <- sdq_eval(data)
data <- data.frame(data,sdq)

```

** Validating the SDQ results

We explore the distribution of SDQ scores for regular and hospital schools separately.The thin grey lines indicate the four scale levels for a *slightly raised* score (15-17), a *high* score (18-19) and a *very high* score (>20).

```{r echo=FALSE}
ggplot(data, aes(x=sdq_total)) +
  geom_vline(aes(xintercept=14.5), color="grey") + 
  geom_vline(aes(xintercept=17.5), color="grey") + 
  geom_vline(aes(xintercept=19.5), color="grey") + 
  geom_histogram(breaks=seq(-.5,40.5,1),alpha=.5) +    
  facet_grid(TYP ~ . ) +   
  xlab("SDQ") + ylab("Number") + 
  theme_grey(base_size=14)  
```

For easy visual judgement, we also plot the empirical cumulative distributions:

```{r echo=FALSE}
ggplot(data, aes(x=sdq_total, color=TYP)) +
  geom_vline(aes(xintercept=14.5), color="grey") + 
  geom_vline(aes(xintercept=17.5), color="grey") + 
  geom_vline(aes(xintercept=19.5), color="grey") + 
  stat_ecdf(geom="step") +
  xlab("SDQ") + ylab("Fraction") + 
  theme_grey(base_size=14)  
```

