---
title: "Problems in school reported by children in regular and hospital schools"
author: "Judith Janschewski (TU Dortmund) and Philipp Berens (Uni Tübingen)"
output:
  pdf_document:
    df_print: kable
    number_sections: yes
  html_notebook: default
---

```{r global_options, include=FALSE}
#library(car)
library(plyr)
library(dplyr)
library(ggplot2)
library(Rmisc)
library(psych)

knitr::opts_chunk$set(fig.width=5,  fig.height=2.5, warning=FALSE, echo=FALSE)
```

# Dataset

```{r echo=FALSE}
# import the data
data <- read.csv("D:/lab/projects/schoolproblems2/data/schooldata_withsdq.csv",na.strings=c("-9","NA"," ","-1","other"))
data <- subset(data, select=c('TYP','SD01','sdq_total','sdq_external','sdq_internal','SS04_01',
                              'SS04_02', 'SS04_03', 'SS04_06', 'SS08_01', 'SS08_02', 'SS08_03', 
                              'SS08_04', 'SS05_05', 'SS03_01', 'SS03_02', 'SS03_03', 'SS03_04',
                              'SB01_01', 'SB01_03', 'SB01_04', 'SB01_02', 'SS02', 'SB04_01',
                              'SD25'))

data <- plyr::rename(data,c("TYP"="school_type","SD01"="gender","SD25"="age"))


data$age <- factor(mapvalues(data$age,from=c(1,2,3,4), to=c("11-12","13-14","15-16","17-18")))

```

The dataset contains questionnaire responses of `r nrow(data)` students from regular and hospital schools. The following tables provides an overview: 

```{r results='asis'}

cdata <- ddply(data, c("gender"), summarise,
               N    = length(sdq_total))

cdata
```


```{r results='asis'}
cdata <- ddply(data, c("school_type"), summarise,
               N    = length(sdq_total))

cdata
```

```{r results='asis'}
cdata <- ddply(data, c("gender", "school_type"), summarise,
               N    = length(sdq_total))

cdata
```


```{r results='asis'}
cdata <- ddply(data, c("age"), summarise,
               N    = length(sdq_total))

cdata
```



# Validating the SDQ results

We explore the distribution of SDQ scores for regular (A) and hospital (K) schools separately. The thin  lines indicate the four scale levels for a *slightly raised* score (15-17), a *high* score (18-19) and a *very high* score (>20).

```{r SDQ_hist, fig.width=6, fig.height=4}
ggplot(filter(data, !is.na(data$gender)), aes(x=sdq_total)) +
  geom_vline(aes(xintercept=14.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=17.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=19.5), color="darkgrey", linetype="dashed") + 
  geom_histogram(aes(y=..density..),breaks=seq(-.5,40.5,1),alpha=.5) +    
  facet_grid(school_type ~ gender ) +   
  xlab("SDQ") + ylab("Number") + 
  theme_minimal(base_size=12)  
```

The  cumulative SDQ distributions of general (A) and hosptial (K) schools show a clear shift between the two student populations:

```{r echo=FALSE, fig.width=4, fig.height=3}
ggplot(data, aes(x=sdq_total, color=school_type)) +
  geom_vline(aes(xintercept=14.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=17.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=19.5), color="darkgrey", linetype="dashed") + 
  stat_ecdf(geom="step") +    
  xlab("SDQ") + ylab("Fraction") + 
  theme_minimal(base_size=12)  
```

The percentage of students with raised SDQ in general schools corresponds approximately to that expected:

```{r fraction_raised, echo=FALSE}

cdata <- ddply(data, c("school_type"), summarise,
                  N    = length(sdq_total),
                  fraction_raised = mean(sdq_total>=15, na.rm=TRUE),
                  fraction_vhigh = mean(sdq_total>=20, na.rm=TRUE))

cdata
```

# Overview of item responses

We first provide an overview of several of the response items.

## Social problems

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS04_01)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I often have conflicts with other students") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS04_02)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I often have conflicts with teachers") + ylab("Relative frequency") + 
theme_minimal(base_size=12)  
```



```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS04_03)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I isolated myself recently in school") + ylab("Relative frequency") + 
theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS04_06)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I do not like large groups like my class") + ylab("Relative frequency") + 
theme_minimal(base_size=12)  
```


```{r , echo=FALSE}

cdata <- ddply(filter(data, !is.na(data$gender)), c("school_type", "gender"), summarise,
                  N    = length(sdq_total),
                  conflict_stud = mean(SS04_01, na.rm=TRUE),
                  conflict_teach = mean(SS04_02, na.rm=TRUE),
                  isolation = mean(SS04_03, na.rm=TRUE),
                  large_groups = mean(SS04_06, na.rm=TRUE))

cdata
```


```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS08_01)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I have friends in my class") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS08_02)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I am bullied by other students") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS08_04)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("Sometimes I am agressive") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```


```{r , echo=FALSE}

cdata <- ddply(filter(data, !is.na(data$gender)), c("school_type", "gender"), summarise,
                  N    = length(sdq_total),
                  friends = mean(SS08_01, na.rm=TRUE),
                  bullying = mean(SS08_02, na.rm=TRUE),
                  outsider = mean(SS08_03, na.rm=TRUE),
                  agressive = mean(SS08_04, na.rm=TRUE))

cdata
```

## School performance

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS03_01)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("My school performance declined last year") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

The next item shows a pretty centered, almost Normal distribution. This is interesting - I would imaginge this may indicate that the students did not understand the question properly.

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS03_02)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I can concentrate well on school content") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```



```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS03_03)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I can work independently") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS03_04)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I quickly loose motivation to work") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r , echo=FALSE}

cdata <- ddply(filter(data, !is.na(data$gender)), c("school_type", "gender"), summarise,
                  N    = length(sdq_total),
                  perf_down = mean(SS03_01, na.rm=TRUE),
                  concentrate = mean(SS03_02, na.rm=TRUE),
                  independent = mean(SS03_03, na.rm=TRUE),
                  loose_mot = mean(SS03_04, na.rm=TRUE))

cdata
```


## Wellbeing/attitude

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SB01_01)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I like to go to school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SB01_03)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I am afraid of going to school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

The next item again may not have been properly understood:

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SB01_04)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("I can easily handle what is asked of me at school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SB01_02)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("School is important for me") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r , echo=FALSE}

cdata <- ddply(filter(data, !is.na(data$gender)), c("school_type", "gender"), summarise,
                  N    = length(sdq_total),
                  like_school = mean(SB01_01, na.rm=TRUE),
                  afraid = mean(SB01_03, na.rm=TRUE),
                  handle = mean(SB01_04, na.rm=TRUE),
                  important = mean(SB01_02, na.rm=TRUE))

cdata
```

## Other items

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SB04_01)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("School is a burden for me") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS02)) +
  geom_histogram(aes(y=..density..,fill=school_type), position = "dodge", binwidth=1, center=1) +    
  facet_grid(. ~gender) +   
  xlab("How often were you absent from school last year?") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r, fig.width=4}

ggplot(filter(data, !is.na(data$gender)), aes(x=SS05_05)) +
  geom_histogram(aes(y=..density..,fill=gender), position = "dodge", binwidth=1, center=1) +    
  xlab("After my stay at the clinic, I want to go back to my old school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```

```{r , echo=FALSE}

cdata <- ddply(filter(data, !is.na(data$gender)), c("school_type", "gender"), summarise,
                  N    = length(sdq_total),
                  burden = mean(SS04_01, na.rm=TRUE),
                  absence = mean(SS02, na.rm=TRUE),
                  goback = mean(SS05_05, na.rm=TRUE))

cdata
```


# Exploratory factor analysis

We use exploratory factor analysis to identify latent factors in the data. We use the psych package. The correlation matrix of all considered items looks like this:

```{r, fig.width=6.6, fig.height=6.6}
sdata <- subset(data, select=c('SS04_01', 'SS04_02', 'SS04_03', 'SS04_06', 'SS08_01', 'SS08_02', 'SS08_03', 
                              'SS08_04', 'SS05_05', 'SS03_01', 'SS03_02', 'SS03_03', 'SS03_04',
                              'SB01_01', 'SB01_03', 'SB01_04', 'SB01_02', 'SS02', 'SB04_01'))

cMat <- cor(sdata, use="pairwise.complete.obs")

corPlot(cMat)
```


```{r, fig.height=10}
solution <- fa(r = cMat, nfactors = 4, rotate = "promax", fm = "pa")
fa.diagram(solution)
```

A provisional interpretation of the factors could be:

* PA1: Affinity to school and a well-being at school

* PA2: Ability to cope with school demands

* PA3: Social integration

* PA4: Aggressive Behavior


## Reliability analysis





