---
title: "Problems in school reported by children in regular and hospital schools"
author: "Judith Janschewski (TU Dortmund) and Philipp Berens (Uni T?bingen)"
output:
  pdf_document:
    df_print: kable
    number_sections: yes
  html_notebook: default
---

```{r global_options, include=FALSE}
#library(car)
library(plyr)
library(dplyr)
library(ggplot2)
library(Rmisc)
library(psych)
library(mice)
library(glmnet)
library(tidyr)

knitr::opts_chunk$set(fig.width=8,  fig.height=2.5, warning=FALSE, echo=FALSE)
```

```{r}
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE2 <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- plyr::rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}
```



# Dataset

```{r echo=FALSE}
# import the data
data <- read.csv("D:/lab/projects/schoolproblems2/data/schooldata_withsdq.csv",na.strings=c("-9","NA"," ","-1","other"))
data <- subset(data, select=c('TYP','SD01','sdq_total','sdq_external','sdq_internal','SS04_01',
                              'SS04_02', 'SS04_03', 'SS04_06', 'SS08_01', 'SS08_02', 'SS08_03', 
                              'SS08_04', 'SS05_05', 'SS03_01', 'SS03_02', 'SS03_03', 'SS03_04',
                              'SB01_01', 'SB01_03', 'SB01_04', 'SB01_02', 'SS02', 'SB04_01',
                              'SD25', 'UB06_01', 'GE06_01', 'GE06_02', 'GE06_03', 'GE06_04', 
                              'GE06_05', 'GE06_06'))

data <- plyr::rename(data,c("TYP"="school_type","SD01"="gender","SD25"="age"))


data$age <- factor(mapvalues(data$age,from=c(1,2,3,4), to=c("11-12","13-14","15-16","17-18")))
data$raised_sdq <- factor(data$sdq_total>=15)
data$group <- factor(data$school_type,levels=c("A","A/SDQ+","K"))
data$group[data$school_type=="A" & data$raised_sdq==TRUE] <- "A/SDQ+" 

fdata <- filter(data, !is.na(data$raised_sdq)) 

```

The dataset contains questionnaire responses of `r nrow(data)` students from regular and hospital schools. The following tables provides an overview: 

```{r results='asis'}

cdata <- ddply(data, c("gender"), summarise,
               N    = length(sdq_total))

cdata
```


```{r results='asis'}
cdata <- ddply(data, c("school_type"), summarise,
               N    = length(sdq_total))

cdata
```

```{r results='asis'}
cdata <- ddply(data, c("gender", "school_type"), summarise,
               N    = length(sdq_total))

cdata
```


```{r results='asis'}
cdata <- ddply(data, c("age"), summarise,
               N    = length(sdq_total))

cdata
```


\newpage
# Validating the SDQ results

We explore the distribution of SDQ scores for regular (A) and hospital (K) schools separately. The thin  lines indicate the four scale levels for a *slightly raised* score (15-17), a *high* score (18-19) and a *very high* score (>20).

```{r SDQ_hist, fig.width=6, fig.height=4}
ggplot(fdata, aes(x=sdq_total)) +
  geom_vline(aes(xintercept=14.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=17.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=19.5), color="darkgrey", linetype="dashed") + 
  geom_histogram(aes(y=..density..),breaks=seq(-.5,40.5,1),alpha=.5) +    
  facet_grid(school_type ~ . ) +   
  xlab("SDQ") + ylab("Number") + 
  theme_minimal(base_size=12)  
```

The  cumulative SDQ distributions of general (A) and hosptial (K) schools show a clear shift between the two student populations:

```{r echo=FALSE, fig.width=4, fig.height=3}
ggplot(data, aes(x=sdq_total, color=school_type)) +
  geom_vline(aes(xintercept=14.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=17.5), color="darkgrey", linetype="dashed") + 
  geom_vline(aes(xintercept=19.5), color="darkgrey", linetype="dashed") + 
  stat_ecdf(geom="step") +    
  xlab("SDQ") + ylab("Fraction") + 
  theme_minimal(base_size=12)  
```

The percentage of students with raised SDQ in general schools corresponds approximately to that expected:

```{r fraction_raised, echo=FALSE}

cdata <- ddply(data, c("school_type"), summarise,
                  N    = length(sdq_total),
                  fraction_raised = mean(sdq_total>=15, na.rm=TRUE),
                  fraction_vhigh = mean(sdq_total>=20, na.rm=TRUE))

cdata
```

\newpage
# Overview of item responses

We first provide an overview of several of the response items.

## Social problems

```{r}
p1 <- ggplot(fdata, aes(x=SS04_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I have conflicts with other students") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS04_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I have conflicts with other students") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)
```

```{r}


p1 <- ggplot(fdata, aes(x=SS04_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I often have conflicts with teachers") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS04_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I often have conflicts with teachers") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)
```



```{r}

p1 <- ggplot(fdata, aes(x=SS04_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I isolated myself recently in school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS04_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I isolated myself recently in schoonl") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r}


p1 <- ggplot(fdata, aes(x=SS04_06)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I do not like large groups like my class") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS04_06)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I am bullied by other students") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```


```{r , echo=FALSE}

cdata <- ddply(fdata, c("group"), summarise,
                  N    = length(sdq_total),
                  conflict_stud = mean(SS04_01, na.rm=TRUE),
                  conflict_teach = mean(SS04_02, na.rm=TRUE),
                  isolation = mean(SS04_03, na.rm=TRUE),
                  large_groups = mean(SS04_06, na.rm=TRUE))

cdata
```


```{r}


p1 <- ggplot(fdata, aes(x=SS08_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I have friends in my class") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS08_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I have friends in my class") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)


```

```{r, fig.width=8, fig.height=3}

p1 <- ggplot(fdata, aes(x=SS08_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I am bullied by other students") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS08_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I am bullied by other students") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r}

p1 <- ggplot(fdata, aes(x=SS08_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I am an outsider") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS08_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I am an outsider") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)
```

```{r}

p1 <- ggplot(fdata, aes(x=SS08_04)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("Sometimes I am agressive") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS08_04)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("Sometimes I am agressive") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)
```


```{r , echo=FALSE}

cdata <- ddply(fdata, c("group"), summarise,
                  N    = length(sdq_total),
                  friends = mean(SS08_01, na.rm=TRUE),
                  bullying = mean(SS08_02, na.rm=TRUE),
                  outsider = mean(SS08_03, na.rm=TRUE),
                  agressive = mean(SS08_04, na.rm=TRUE))

cdata
```

## School performance

```{r}

p1 <- ggplot(fdata, aes(x=SS03_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("My school performance declined last year") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS03_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("My school performance declined last year") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```


```{r}

p1 <- ggplot(fdata, aes(x=SS03_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I can concentrate well in school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS03_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I can concentrate well in school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)
```



```{r}

p1 <- ggplot(fdata, aes(x=SS03_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I can work independently") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS03_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I can work independently") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r}

p1 <- ggplot(fdata, aes(x=SS03_04)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I qickly loose motivation to work") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS03_04)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I quickly loose motivation to work") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r , echo=FALSE}

cdata <- ddply(fdata, c("group"), summarise,
                  N    = length(sdq_total),
                  perf_down = mean(SS03_01, na.rm=TRUE),
                  concentrate = mean(SS03_02, na.rm=TRUE),
                  independent = mean(SS03_03, na.rm=TRUE),
                  loose_mot = mean(SS03_04, na.rm=TRUE))

cdata
```


## Wellbeing/attitude

```{r}

p1 <- ggplot(fdata, aes(x=SB01_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I like to go to school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SB01_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I like to go to school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r}

p1 <- ggplot(fdata, aes(x=SB01_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I am afraid of going to school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SB01_03)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I am afraid of going to school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r}

p1 <- ggplot(fdata, aes(x=SB01_04)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("I can easily handle what\n is asked of me at school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SB01_04)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("I can easily handle what\n is asked of me at school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)


```

```{r}


p1 <- ggplot(fdata, aes(x=SB01_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("School is important for me") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SB01_02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("School is important for me") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)


```

```{r , echo=FALSE}

cdata <- ddply(fdata, c("group"), summarise,
                  N    = length(sdq_total),
                  like_school = mean(SB01_01, na.rm=TRUE),
                  afraid = mean(SB01_03, na.rm=TRUE),
                  handle = mean(SB01_04, na.rm=TRUE),
                  important = mean(SB01_02, na.rm=TRUE))

cdata
```

## Other items

```{r}

p1 <- ggplot(fdata, aes(x=SB04_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("School is a burden for me") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SB04_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("School is a burden for me") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r}

p1 <- ggplot(fdata, aes(x=SS02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=1, center=1) +    
  xlab("How often were you abesent from school last year") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

p2 <- ggplot(fdata, aes(x=SS02)) +
  geom_histogram(aes(y=..density..,fill=group), position = "fill", binwidth=1, center=1) +    
  xlab("How often were you absent from school last year") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  +
  scale_fill_brewer(palette="Spectral")

multiplot(p1,p2,cols=2)

```

```{r, fig.width=4}

ggplot(fdata, aes(x=SS05_05)) +
  geom_histogram(aes(y=..density..,fill=gender), position = "dodge", binwidth=1, center=1) +    
  xlab("After my stay at the clinic, \n I want to go back to my old school") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  


```

```{r , echo=FALSE}

cdata <- ddply(fdata, c("group"), summarise,
                  N    = length(sdq_total),
                  burden = mean(SS04_01, na.rm=TRUE),
                  absence = mean(SS02, na.rm=TRUE),
                  goback = mean(SS05_05, na.rm=TRUE))

cdata
```


\newpage
# Classification between healthy students and students with psychological problems

We have quite a few missing values in the predictor variables. For now, we impute them using the mice package [link](https://www.jstatsoft.org/article/view/v045i03).

```{r}

x <- select(data,c(-gender,-school_type,-age,-sdq_total,-sdq_internal,-sdq_external,-SS05_05,-raised_sdq,-group)) 

xi <- mice(x, printFlag = FALSE, seed = 1000)
xi <- mice::complete(xi,1) %>%
      as.matrix() 

```

We try to predict whether a student is either a student of a regular school with an SDQ<15 or a student of a hospital school/student of a regular school with SDQ>15 using the glmnet package [link](http://web.stanford.edu/~hastie/glmnet/glmnet_alpha.html).

```{r}

# format data
y <- as.matrix(data$group=="K" | data$group=="A/SDQ+")   # labels

idx <- !is.na(y)
xx <- xi[idx,]
y <- factor(y[idx,])
g <- data$group[idx]

# fit models
cvfit = cv.glmnet(xx, y, family="binomial", nfolds = 50, alpha = .95)
w <- coef(cvfit, s = "lambda.1se") %>% as.matrix()
yp <- predict(cvfit, newx = xx, s = "lambda.1se")

res <- data.frame(y,g,yp)
colnames(res) <- c('y','group','yp')

```

The performance of this classifier is:

```{r}
res$pred <- res$yp>0
res$correct <- res$pred == res$y

cdata <- ddply(res, c("group"), summarise,
                  N    = length(correct),
                  p_correct = mean(correct, na.rm=TRUE) * 100)

cdata
```

We can plot the weights to visualize influential items:

```{r, echo=FALSE, fig.width=6,fig.height=6}

# collect weights
weights <- data.frame(rownames(w)[2:length(w)],w[2:length(w)])
colnames(weights) <- c('item','weight')

p1 <- ggplot(weights, aes(x=item, y= weight)) + 
   geom_bar(stat = "identity", position = "dodge") +
   theme_minimal(base_size=12) +
   theme(axis.text.x = element_text(angle=45, vjust=.5))

p2 <- ggplot(res,aes(x=yp,fill=group)) +
    geom_histogram(position="dodge") +
   theme_minimal(base_size=12) +
  xlab('Projection on w')

multiplot(p1,p2,cols=1)

```

The influential items are :

* School is important for me

* School is a burden for me

* I have conflicts with teachers

\newpage
# SDQ Prediction

We have quite a few missing values in the predictor variables. For now, we impute them using the mice package [link](https://www.jstatsoft.org/article/view/v045i03).

```{r}

x <- select(data,c(-gender,-school_type,-age,-sdq_total,-sdq_internal,-sdq_external,-SS05_05,-raised_sdq,-group)) 

xi <- mice(x, printFlag = FALSE, seed = 1000)
xi <- mice::complete(xi,1) %>%
      as.matrix() 

```

We try to predict total, internal and external SDQ score from the response items using the glmnet package [link](http://web.stanford.edu/~hastie/glmnet/glmnet_alpha.html).

```{r}

# format data
y <- as.matrix(data$sdq_total)   # labels
ye <- as.matrix(data$sdq_external)   # labels
yi <- as.matrix(data$sdq_internal)   # labels

idx <- !is.na(y)
xi <- xi[idx,]
y <- y[idx,]
ye <- ye[idx,]
yi <- yi[idx,]

# fit models
cvfit = cv.glmnet(xi, y, type.measure = "mse", nfolds = 50, alpha = 1)
w <- coef(cvfit, s = "lambda.1se") %>% as.matrix()
yp <- predict(cvfit, newx = xi, s = "lambda.1se")

cvfit_e = cv.glmnet(xi, ye, type.measure = "mse", nfolds = 50, alpha = 1)
we <- coef(cvfit_e, s = "lambda.1se") %>% as.matrix()
yep <- predict(cvfit_e, newx = xi, s = "lambda.1se")

cvfit_i = cv.glmnet(xi, yi, type.measure = "mse", nfolds = 50, alpha = 1)
wi <- coef(cvfit_i, s = "lambda.1se") %>% as.matrix()
yip <- predict(cvfit_i, newx = xi, s = "lambda.1se")

```

We can plot the weights to visualize influential items:

```{r, echo=FALSE, fig.width=7,fig.height=5}

# collect weights
weights <- data.frame(rownames(w)[2:length(w)],w[2:length(we)],we[2:length(we)],wi[2:length(we)])
colnames(weights) <- c('item','total','external','internal')
weights <- gather(weights, sdq, weight, total:internal)

# sort by sdq interal weight
ord <- sort.int(abs(weights$weight[weights$sdq=="internal"]), index.return=TRUE)$ix
weights$item <- factor(weights$item, levels = weights$item[ord])

ggplot(weights, aes(x=item, y= weight)) + 
   geom_bar(stat = "identity", aes(fill=sdq), position = "dodge") +
   theme_minimal(base_size=12) +
   theme(axis.text.x = element_text(angle=45, vjust=.5))

```

This indicates that the items predictive for a high score of the SDQ internalizing scale are disjunct from those predictive for a high score on the externalizing SDQ scale.

Predictive of a high score on the internalizing SDQ scale are (in order of decreasing weight):

* I am afraid to go to school
* I isolated myself recently
* School is a burden for me
* I don't like large groups like my class
* I have friends in my class
* I am an outsider 
* I am bullied by other students

Predictive of a high score an the externalizing SDQ scale are:

* I have conflicts with teachers
* I can work independently
* My school performance dropped during the last year
* I have conflicts with other students

Interesting is "I can easily handle what is asked of me at school" - it is not strongly predictive for either internalizing/externalizing, but highly predictive of the total score. Most other items that are preditive for one are also predictive for the total.

```{r, echo=FALSE, fig.height=3, fig.width=6}

preds = data.frame(c(y,ye,yi), 
                   c(yp,yep,yip),
                   c(rep('total',length(y)),rep('external',length(y)),rep('internal',length(y))),
                   stringsAsFactors = TRUE) 
colnames(preds) <- c('orig','pred','sdq')

ggplot(preds, aes(x=orig,y=pred)) + 
  geom_point(size=.75, position = position_jitter(width=.1)) +
  facet_grid(.~sdq) +
  theme_minimal()
  

```

The correlation between the original values and the predictions is high (total: `r cor(y,yp)`, internal: `r cor(yi,yip)`, external: `r cor(ye,yep)`), indicating good prediction.


\newpage
# Support through teachers

```{r , echo=FALSE}

cdata <- ddply(fdata, c("group"), summarise,
                  N    = length(sdq_total),
                  nans = sum(is.na(UB06_01)),
                  fraction = mean(is.na(UB06_01)))

cdata
```


```{r}

ggplot(data, aes(x=UB06_01)) +
  geom_histogram(aes(y=..density..,fill=group), position = "dodge", binwidth=10, center=5) +     
  xlab("- Support for problems +") + ylab("Relative frequency") + 
  theme_minimal(base_size=12)  
```



```{r}
ggplot(data, aes(x=sdq_total,y=UB06_01, color=group)) +
  geom_point(position=position_jitter(width=.3), size=1) + 
  geom_smooth(method=lm)

```

The spearman correlation is `r cor(data['sdq_total'],data['UB06_01'],use="pairwise.complete.obs", method="spearman")` for total SDQ, `r cor(data['sdq_internal'],data['UB06_01'],use="pairwise.complete.obs", method="spearman")` for internalizing SDQ and `r cor(data['sdq_external'],data['UB06_01'],use="pairwise.complete.obs", method="spearman")` for externalizing SDQ. 


```{r}
mdl <- lm(UB06_01 ~ sdq_total + school_type,data=data)
summary(mdl)

```


```{r}
mdl2 <- lm(UB06_01 ~ sdq_internal * sdq_external + school_type,data=data)
summary(mdl)

```


```{r, fig.width=3,fig.height=4}

tmp <- summarySE2(data, measurevar = "UB06_01",groupvars = "SS05_05", na.rm=TRUE)

ggplot(tmp, aes(x=SS05_05, y=UB06_01)) + 
    geom_errorbar(aes(ymin=UB06_01-se, ymax=UB06_01+se), width=.1) +
    geom_point() +    
  xlab("After my stay at the clinic, \n I want to go back to my old school") + ylab("I feel supported") +
  ylim(c(0,100))+ xlim(c(0,6))+
  theme_minimal(base_size=12)  
```



#I want to go back to school vs I feel supported by my teachers

#Predict I feel supported from positive and negative properties

\newpage
# Exploratory factor analysis

We use exploratory factor analysis to identify latent factors in the data. We use the psych package. The correlation matrix of all considered items looks like this:

```{r, fig.width=6.6, fig.height=6.6}
sdata <- subset(data, select=c('SS04_01', 'SS04_02', 'SS04_03', 'SS04_06', 'SS08_01', 'SS08_02', 'SS08_03', 
                              'SS08_04', 'SS05_05', 'SS03_01', 'SS03_02', 'SS03_03', 'SS03_04',
                              'SB01_01', 'SB01_03', 'SB01_04', 'SB01_02', 'SS02', 'SB04_01'))

cMat <- cor(sdata, use="pairwise.complete.obs")

solution <- fa(r = cMat, nfactors = 4, rotate = "promax", fm = "pa")
fa.diagram(solution)
```

A provisional interpretation of the factors could be:

* PA1: Affinity to school and a well-being at school

* PA2: Ability to cope with school demands

* PA3: Social integration

* PA4: Aggressive Behavior







